# Immune Cell Infiltration {#immune_infiltrate}
In previous chapters, many genes responsible for immune/inflammotary response showed a trend of elevated expression levels in FSHD samples. This chapter made further analysis on examing the evidence of immune cell infiltration in FSHD affected muscle biopsies and the associtation with disease activities and progression. 

```{r load-data-chap7, include=FALSE}
library(tidyverse)
library(DESeq2)
library(xlsx)
library(pheatmap)
pkg_dir <- "/fh/fast/tapscott_s/CompBio/RNA-Seq/hg38.FSHD.biopsy.all"
source(file.path(pkg_dir, "scripts", "manuscript_tools.R"))
load(file.path(pkg_dir, "public_data", "cluster_df.rda"))
load(file.path(pkg_dir, "public_data", "sanitized.dds.rda"))
load(file.path(pkg_dir, "public_data", "sanitized.rlg.rda"))
load(file.path(pkg_dir, "public_data", "deseq2_design_cluster.rda"))
sanitized.dds$RNA_cluster_5 <- cluster_df$RNA_cluster
sanitized.dds$cluster_color <- cluster_df$color
sanitized.dds$new_cluster_name <- ifelse(cluster_df$RNA_cluster == "A_Cntr", "Control",
                                         as.character(cluster_df$new_cluster_name))
sanitized.dds$new_cluster_name <- factor(sanitized.dds$new_cluster_name, 
                                         levels=c("Control", "Mild", "Moderate", 
                                                  "IG-High", "High", "Muscle-Low"))
sanitized.rlg$new_cluster_name <- sanitized.dds$new_cluster_name
```

## Immune cell markers
The immune cell markers we considered here are annotated by Nano String. There are total 24 cell types with 105 genes.

```{r load_immune_cell_info}
immune_cell <- read.delim(file.path(pkg_dir, "extdata", "immune_cell_markers", 
                                    "nanoStringImmuneCellMarkers.txt"))
immune_cell <- immune_cell %>% 
  dplyr::filter(gene %in% rowData(sanitized.dds)$gene_name) %>%
  dplyr::mutate(gencode_id = get_ensembl(as.character(gene), sanitized.dds))
#' list the genes
tmp <- immune_cell %>% 
  group_by(cellType) %>%
  summarise(markers=paste(as.character(gene), collapse=", "))
knitr::kable(tmp, caption="Immune cell makers by cell types.")
```

## Immune cell type scores by TPM per Samples
For each sample, 24 immune cell type scores are given by average $\log_{10}(TPM+1)$ of associated markers. Boxplot below presents immune cell type scores of each samples, seperated by FSHD classes.

```{r get_score_TPM}
lv <- levels(immune_cell$cellType)
immune_cell_tpm <- sapply(lv, function(cell_type) {
   immune_id <- immune_cell %>% dplyr::filter(cellType == cell_type) %>%
      pull(gencode_id)
   sub <- sanitized.dds[immune_id]
   tpm_score <- colMeans(log10(assays(sub)[["TPM"]]+1))
})

immune_cell_tpm <- as.data.frame(immune_cell_tpm) %>%
  rownames_to_column(var="sample_name") %>%
  add_column(cluster = sanitized.dds$new_cluster_name, 
             cluster_color =sanitized.dds$cluster_color) %>%
  gather(key=cell_type, value=mean_tpm, -sample_name, -cluster, -cluster_color) 

cluster_tpm_score <- immune_cell_tpm %>%
  group_by(cell_type, cluster) %>%
  summarise(cell_type_per_cluster=mean(mean_tpm))
```

```{r vis_infiltrate, fig.height=12, fig.width=10}
color_manual <- immune_cell_tpm %>% group_by(cluster) %>% 
  summarise(color=unique(cluster_color)) 

ggplot(immune_cell_tpm, aes(x=cluster, y=mean_tpm, color=cluster)) +
  geom_boxplot() +
  facet_wrap(~ cell_type, scale="free") +
  theme_minimal() +
  theme(axis.text.x=element_blank(), legend.justification=c(1,0), legend.position=c(1,0)) +
  scale_color_manual(values=color_manual$color) +
  labs(y="Avg. log10(TPM +1)", x="FSHD classes", title="Immune cell infiltrate score by TPM")
```

## Immune cell marker scores by logFC per FSHD classes
For each immune cell markers, scores are given as the logFC for each FSHD classes (Mild, Moderate, IG-High, High and Muscle-Low compared to the controls).

```{r immune_marker_logFC, warning=FALSE}
names <- resultsNames(deseq2_design_cluster)[-1]
immune_marker_logFC <- lapply(resultsNames(deseq2_design_cluster)[-1], function(results_name){
  res <- results(deseq2_design_cluster, name=results_name)
  res[as.character(immune_cell$gencode_id), "log2FoldChange", drop=FALSE]
})
names(immune_marker_logFC) <- c("Mild", "Moderate", "IG-High", "High", "Muscle-Low")
immune_marker_logFC <- as(do.call(cbind, immune_marker_logFC), "data.frame") %>%
  rownames_to_column(var="gencode_id") %>%
  left_join(immune_cell, by="gencode_id")
```

```{r viz_immune_marker_logFC, fig.height=10, fig.width=10}
#' boxplot of immune marker per celltype by FSHD classes
tidy_marker_logFC <- immune_marker_logFC %>%
  gather(key=classes, value=log2FoldChange, -gencode_id, -gene, -altName, -cellType) %>%
  dplyr::mutate(classes=factor(classes, 
                               levels=paste0(c("Mild", "Moderate", "IG-High",
                                               "High", "Muscle-Low"), ".log2FoldChange")))
ggplot(tidy_marker_logFC, aes(x=classes, y=log2FoldChange, color=classes)) +
  geom_boxplot(outlier.shape=NA) +
  geom_jitter(size=0.5, width=0.5) +
  facet_wrap(~cellType, scale="free") +
  scale_color_manual(values=color_manual$color[-1]) +
  theme_minimal() +
  theme(axis.text.x=element_blank(), legend.justification=c(1,0), legend.position=c(1,0)) 
```

## Frequency of significance 
Code chunk below calculates frequency of significent markers (log2FC > 2) by cell types.

```{r freq_of_significance, fig.height=8, fig.width=11}
freq_sig <- tidy_marker_logFC %>%
  group_by(cellType, classes) %>%
  summarise(n_markers=n(),
            frequency = sum(log2FoldChange >2) / n_markers) %>%
  dplyr::mutate(label = paste0(cellType, " (", n_markers, ")"))
  #summarise(n_markers=nrow(.)/length(levels(classes)), 
  #          freq = sum(log2FoldChange > 2) / n_markers)
ggplot(freq_sig, aes(x=classes, y=frequency, color=classes, fill=classes)) +
  geom_bar(stat="identity") +
  facet_wrap(~ label) +
  theme_minimal() +
  scale_color_manual(values=color_manual$color[-1]) +
  scale_fill_manual(values=color_manual$color[-1]) +
  theme(axis.text.x=element_blank(), legend.justification=c(1,0), legend.position=c(1,0),
        legend.key.size=unit(.5, "cm")) 
  
```

```{r avg_rlog_by_samples, fig.height=12, fig.width=10, eval=FALSE, include=FALSE}
# gapped by classes and cell type
sub <- sanitized.rlg[as.character(immune_cell$gencode_id)]
immune_rlog <- assay(sub)
rownames(immune_rlog) <- rowData(sub)$gene_name
anno_row <- immune_cell %>% dplyr::select(gene, cellType) %>%
  column_to_rownames(var="gene")
anno_col <- as(colData(sub)[, c("pheno_type", "new_cluster_name"), drop=FALSE], "data.frame")
pheatmap(immune_rlog,
         scale="row", 
         annotation_col = anno_col,
         annotation_row = anno_row,
         fontsize_col = 6,
         fontsize_row = 6
         )
```

## Average rlog per sample

```{r boxplot_ave_rlog_per_sample}
sub <- sanitized.rlg[as.character(immune_cell$gencode_id)]
immune_sample_score <- data.frame(average=colMeans(assay(sub))) %>%
  rownames_to_column(var="sample_name") %>%
  add_column(pheno_type = sub$pheno_type,
             classes = sub$new_cluster_name)
ggplot(immune_sample_score, aes(x=classes, y=average, color=classes)) +
  geom_boxplot(outlier.shape=NA) +
  geom_jitter(width=0.3) +
  theme_minimal() +
  scale_color_manual(values=color_manual$color)
```