# Biopsy Quality Control {#biopsy_qc}
Blood, fat, muscle markers and scored by $\log_{10}(TPM + 1)$. 

```{r load-data-chap9, include=FALSE}
library(tidyverse)
library(DESeq2)
library(xlsx)
library(pheatmap)
pkg_dir <- "/fh/fast/tapscott_s/CompBio/RNA-Seq/hg38.FSHD.biopsy.all"
source(file.path(pkg_dir, "scripts", "manuscript_tools.R"))
load(file.path(pkg_dir, "public_data", "cluster_df.rda"))
load(file.path(pkg_dir, "public_data", "sanitized.dds.rda"))
sanitized.dds$cluster_color <- cluster_df$color
sanitized.dds$new_cluster_name <- ifelse(cluster_df$RNA_cluster == "A_Cntr", "Control",
                                         as.character(cluster_df$new_cluster_name))
sanitized.dds$new_cluster_name <- factor(sanitized.dds$new_cluster_name, 
                                         levels=c("Control", "Mild", "Moderate", 
                                                  "IG-High", "High", "Muscle-Low"))
```

## Blood, fat and muscle cell markers
```{r define_markers}
markers <- tibble(cell_type=c(rep("blood", 3), rep("fat", 3), rep("muscle", 3)),
                  gene_name=c("HBA1", "HBA2", "HBB",
                              "FASN", "LEP", "SCD",
                              "ACTA1", "TNNT3", "MYH1"))  %>%
  dplyr::mutate(gencode_id = get_ensembl(gene_name, rse=sanitized.dds)) %>%
  dplyr::mutate(cell_type = factor(cell_type))
knitr::kable(markers)
```


## Define score
For each cell type, the marker score for each sample is $\frac{1}{n}\Sigma_{i=1}^{n}\log_{10}(TPM_i + 1)$, where $i$ denotes genes and $n$ is the number of genes of the cell type.

```{r define_score}
# average cell type log10(TPM+1)
lv <- levels(markers$cell_type)
celltype_tpm <- sapply(lv, function(type) {
   id <- markers %>% dplyr::filter(cell_type == type) %>%
      pull(gencode_id)
   sub <- sanitized.dds[id]
   tpm_score <- colMeans(log10(assays(sub)[["TPM"]]+1))
})

celltype_tpm <- as.data.frame(celltype_tpm) %>%
  rownames_to_column(var="sample_name") %>%
  add_column(pheno_type = sanitized.dds$pheno_type,
             classes = sanitized.dds$new_cluster_name,
             classes_color = sanitized.dds$cluster_color) 
```
```{r fat_tpm, echo=FALSE, eval=FALSE}
# this chunk is to generate an excel sheet 
sample_tpm <- lapply(lv, function(type){
  sub_marker <- markers %>% dplyr::filter(cell_type == type) 
  id <- sub_marker %>% pull(gencode_id)
  names(id) <- pull(sub_marker, "gene_name")
  sub <- sanitized.dds[id]
  
  df <- as.data.frame(t(assays(sub)[["TPM"]]+1)) %>%
     dplyr::rename(!!id) %>%
      rownames_to_column(var="sample_name") %>%
     add_column(pheno_type=sanitized.dds$pheno_type,
                new_cluster_name=sanitized.dds$new_cluster_name)
}) 
names(sample_tpm) <- lv
#' write an excel sheet
#muscle_logTPM <- sample_tpm[["muscle"]] %>% dplyr::filter(new_cluster_name == "Muscle-Low")
muscle_TPM <- sample_tpm[["muscle"]] %>% 
  arrange(new_cluster_name) %>%
  dplyr::mutate(ACTA1.pass = ifelse(ACTA1 > 200, "Yes", "NO"),
                TNNT3.pass = ifelse(TNNT3 > 50, "Yes", "NO"),
                MYH1.pass = ifelse(MYH1 > 10, "Yes", "NO"))
library(xlsx)
write.xlsx(muscle_TPM, file=file.path(pkg_dir, "extdata",
                                      "Wellstone_RNAseq_muscle_content.xlsx"),
           sheetName="Muscle content (TPM)",
           row.names=FALSE)
```

## Visualization
The blood, fat and muscle content scores are given to each sample. Boxplot below illustrates that the FSHD biopsies categorized in the Muscle-Low class are fatter and "blooder" and have less muscle content than the rest of the biopsies.

```{r boxplot_for_qc}
tmp <- celltype_tpm %>% gather(key=cell_type, value=score, -sample_name,
                               -pheno_type, -classes, -classes_color) %>%
  dplyr::mutate(cell_type = factor(cell_type))

color_manual <- celltype_tpm %>% group_by(classes) %>% 
  summarise(color=unique(classes_color)) 

ggplot(tmp, aes(x=classes, y=score, color=classes)) +
  geom_boxplot() +
  facet_wrap(~ cell_type, scale="free",  nrow=2) +
  theme_minimal() +
  theme(axis.text.x=element_blank(), legend.justification=c(1,0), legend.position=c(1,0)) +
  scale_color_manual(values=color_manual$color) +
  labs(y="Avg. log10(TPM +1)", x="FSHD classes", 
       title="Blood, fat, muscle content score by TPM")
```



## Determine "outliers"
```{r build_distribution}
#' use TPM or use log10(TPM+10)?
markers_dds <- sanitized.dds[markers$gencode_id]
tpm <- t(assays(markers_dds)[["TPM"]])
tpm_log <- log10(tpm+1)

#' get distribution of test statistics by bootstrapping
library(modelr)
get_test_stats <- function(expr, n=500) {
  set.seed(123)
  boot <- bootstrap(as.data.frame(expr), n)
  models <- map(boot$strap, function(x) {
    colMeans(as.data.frame(x))
  })
  #' row bind
  do.call(rbind, models)
}

#' 1) get distribution
set.seed(123)
test_stats <- get_test_stats(tpm, n=1000)
tmp <- as.data.frame(test_stats) %>% gather(key=markers, value=statistics)
gg <- ggplot(tmp, aes(statistics)) +
  geom_density() +
  facet_wrap(~ markers, scale="free") +
  theme_minimal()
gg
```
```{r get_ci}
#' 2) get confidence interval
ci <- as.data.frame(test_stats) %>%
  gather(key=marker, value=statistics) %>%
  group_by(marker) %>%
  summarise(ci = list(mean_cl_normal(statistics, conf.int=0.99, na.rm=TRUE))) %>% 
                        #rename(mean=y, lwr=ymin, upr=ymax))) %>% 
  unnest

 stats <- as.data.frame(test_stats) %>%
  gather(key=gencode_id, value= statistics) %>% 
  group_by(gencode_id) %>%
  summarise(mu = mean(statistics, na.rm = TRUE),
            ssd = sd(statistics, na.rm = TRUE),
            bootstrap_n=n()) %>%
   left_join(markers, by="gencode_id") %>%
   arrange(cell_type) %>%
   dplyr::mutate(min= mu - 5*ssd, max=mu + 5*ssd)
 ggplot(stats, aes(x=mu, y=gene_name)) +
   geom_errorbarh(aes(xmin=mu-3*ssd, xmax=mu+3*ssd), height=0.2) +
   facet_wrap( ~cell_type, scale="free", ncol=1) +
   theme_bw() +
   labs(x="mu +/- 3 sigma")
```
Outliers: outside of $\mu \pm 3\sigma$?
```{r determine_outliers}
stats %>% dplyr::filter(cell_type=="muscle")

a <- as.data.frame(tpm ) %>%
  rownames_to_column(var="sample_name") %>%
  add_column(cluster=sanitized.dds$new_cluster_name) %>%
  dplyr::filter(cluster == "Muscle-Low") %>%
  dplyr::select(sample_name, ENSG00000109061.9, ENSG00000130595.16, ENSG00000143632.14)

flag <- lapply(colnames(tpm), function(gene_id) {
  lower <- stats %>% dplyr::filter(gencode_id == gene_id) %>% pull(min)
  upper <- stats %>% dplyr::filter(gencode_id == gene_id) %>% pull(max)
  
  x <- tpm_log[, gene_id]
  flag <- rep(0, length(x))
  flag[x < lower] <- -1
  flag[x > upper] <- 1
  flag
})
names(flag) <- colnames(tpm_log)
flag <- do.call(cbind, flag)
rownames(flag) <- rownames(tpm_log)

flag <- as.data.frame(flag) %>%
  rownames_to_column(var="sample_name") %>%
  add_column(cluster=sanitized.dds$new_cluster_name)

flag %>% dplyr::filter(cluster=="Muscle-Low")
```

use permutation test? calculate the p-value by constructing mean distribution?


## Determine "bad" biopsies?

## Sample 01-0037
``

## Tables
```{r kable_celltype_tpm}
tb <- dplyr::select(celltype_tpm, -classes_color) %>%
  arrange(classes)
knitr::kable(tb, caption="Blood, fat and muscle content index.")
#' blood, fat and muscle mean score for each class
tb2 <- tb %>%
  group_by(classes) %>%
  summarise(blood_mean = mean(blood),
            fat_mean = mean(fat),
            muscle_mean = mean(muscle))
knitr::kable(tb2)
```
